An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging

### 2018-02-19 HowtoHowtoCapon/document.tex
Parallel build starting to work
2_MVDR ok. 3_LCA inlcuded

\title{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}

\author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
        Jon~Petter~\AA{}Åsen,~\IEEEmembership{Student Member,~IEEE,} %
        Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
        Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, 0316 Oslo, Norway.}%
\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, 7491 Trondheim, Norway.}% <-this % stops a space
% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
}

% The paper headers
\markboth{IEEE Journal of Oceanic Engineering}%
{A Robust and Optimized GPU Implementation of the MVDR beamformer}

% Publishers ID mark:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
% Two limiting factors of modern phased array sonar imaging systems are the angular resolution and contrast. Adaptive beamformers target these limitatitons by weighting sensors dynamically to fit the impinging wavefield.

The minimum variance distortionless response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging. Unfortunately, it is very computationally complex because a spatial covariance matrix must be estimated and inverted for each image pixel. This may discourage its use unnecessarily in sonar systems which are continuously being pushed to ever higher imaging ranges and resolutions.

@@

% We benchmarked optimized implementations of the MV beamformer in MATLAB and C, running on a quad-core Intel Xeon E5620, and in CUDA running on a high-end nVidia card. The beamformers operated on data from the 32x4 element Kongsberg Martime HISAS1030 sonar, using a conservative set of robustification parameters. 
%The CPU would then be free to take on the rest of the beamforming as well.
\end{abstract}

% Keywords (normally not used for peer reviews)
\begin{IEEEkeywords}
Adaptive beamforming, MVDR, Capon, sonar, active sonar, GPU, graphics processing unit, complexity.
\end{IEEEkeywords}
}

@@

\subsection{Beamforming}

\ifPeerReview
\begin{figure}[H]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/beamforming.eps}
\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

@@

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}). This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays by
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
\end{align}
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/implementation.eps}
\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth:\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed; \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver;\newline
3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
\end{figure}
As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{finalR});
\item computing $\eRi\1$ in (\ref{weights});
\item computing the beamformer output $z$ in (\ref{finalZ}).
\end{enumerate}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

@@

If we neglect spatial and temporal averaging, then the computation of $\eR$ is reduced to a single outer product with complexity of O($M^2$), and the inversion is then of O($M^3$). This might lead us to believe that the inversion step is by far the most complex. But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$). Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske3\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_complexity.eps}
\fi%
\caption{Per-pixel computational complexity of the steps in MVDR beamforming (before any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
\end{figure}
To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}). The only excluded operation was partial pivoting used in the inversion step, but this should contribute marginally to the end result. The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$. The results are shown in \Fig{mvdr_complexity}.

@@

\ifPeerReview
\begin{figure}[t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske4\figPostfix.eps}
\else
\begin{figure}[t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_build_R.eps}
\fi%
\caption{Step 1: Building $\eR$. This is a visualization of how $\eR$ could be built in a case with $M=5$ sensors, with subarray size $L=3$ and temporal averaging set to $K=1$. Here $\R'_{l}$ is the sample covariance matrix for the $l$th subarray, and $\breve{\R}$ is the average of $N_K$ of these. Note that instead of performing the temporal sum last as here, one could take more temporal samples into consideration in the computation of each $r_{ij}$.}\label{mvdr_build_R}
\end{figure}

@@

\subsection{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}\label{computing_eR}

@@

In Fig. \ref{mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may:
%
\begin{enumerate}
\item exploit the fact that $\eR$ is Hermitian positive semidefinite to compute only one half of it;
\item avoid redundant multiplications;
\item avoid redundant summations by first computing the upper row of $\eR$ and then adding/subtracting to these iteratively to find the diagonals of $\eR$.
\end{enumerate}
%
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske5\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
\fi%
\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig. \ref{mvdr_complexity} (higher is better) . Note how the arithmetic count is reduced by a factor 4-10 in the memory-optimized case, and by a factor 6-22 in the instruction-optimized case.}\label{mvdr_complexity_optimized}
\end{figure}

@@

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske6\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
\fi%
\caption{MVDR implementated on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \emph{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes interthread communication. \emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}. \emph{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sums all the channels up.}\label{mvdr_implementation}
\end{figure}

@@

\subsection{Solving $\hat{\boldsymbol{R}}$$\;\!^{-1}$$\mathbf{1}$}

@@

\subsection{Computing $z$}

@@

The GPU design is summarized and compared to our CPU designs in table \ref{implementation_summary}.
\begin{table}[!h]\centering%\normalsize
% \begin{tabular}[c]{l r r r@{}  l}\hline
\begin{tabular}[l]{l >{\raggedright}p{.27\linewidth} p{.28\linewidth}}\hline
\rowcolor{tabBlue} & \bf GPU & \bf CPU \\\hline
Pixel processing & In batches & One by one \\
Threads per pixel & Multiple & One \\
1. Building $\eR$:  & & \\
\ \ \ Values computed & All & Only upper triangular \\
\ \ \ Optimization    & Minimized memory consumption & Minimized number of instructions \\
2. Computing $\eRi\1$: & & \\
\ \ \ Method & Nvidia's batch\newline Gauss Jordan solver & Custom inline\newline Cholesky solver \\
Cache utilization & Efficient & Not efficient \\
Data transfers & Minimum & Moderate
\end{tabular}
\caption{Implementation summary.}\label{implementation_summary}
\end{table}

@@

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske7\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
\fi%
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. The image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP). Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}

@@

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske8\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark.eps}
\fi%
\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in Fig. \ref{mvdr_complexity}.\newline
\emph{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
\end{figure}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske9\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/code_assessment.eps}
\fi%
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1 megapixel (MP) image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske10\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/code_assess_flops.eps}
\fi
\caption{\protect Code efficiency. An estimate of the number of floating point operations per second (Flop/s), found by dividing the theoretical complexity curves by actual run-times. This is a crude measure as it does not include any other instructions than the actual arithmetic operations in the MVDR computation.}\label{code_assess_flops}
\end{figure}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske11\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark_boston.eps}
\fi
\caption{\protect MVDR benchmarks from Boston HPC centre with the new high-end Nvidia K20 Kepler GPU. The exact same scenario and code as in Fig. \ref{benchmarks} was used here. With no code alterations the performance was only improved marginally compared to running on the Quadro 6000.}\label{benchmarks_boston}
\end{figure}

@@

Even with all our efforts we were only able to obtain processing rates of 40\;GFlop/s on average in the desirable range of subarray sizes (\Fig{mvdr_complexity}). This is mainly due to the inversion step, since for the most part we obtained more than 100\;GFlop/s when building $\eR$. While these numbers are slightly underestimated, they are regardless far away from the Quadro 6000's peak of 1\;TFlop/s. Again, we believe this is due to the design being bound by memory bandwidth. This belief is further supported by the test results from running the code on the new K20 Kepler card, where we saw a modest factor 1.5-2 speedup. Although the Kepler card peaks at 3.5-4\;Tflop/s, its shared memory bandwidth is only approximately twice that of the Quadro 6000, hence matches the observed speedup. However, it is likely that the Kepler card would perform better had we optimized our design for it. For scientific computing, the Kepler boards are considered more attractive, with the introduction of features such as kernels calling kernels, more registers, more shared memory, and better mechanisms for hiding or removing data transfers.

@@

\appendices

@@

\appendices

@@

% \bibliographystyle{IEEEtran}
% \bibliography{library.bib}

@@

\ifPhdDoc\else
\input{bio/jo_inge_buskenes}
\input{bio/jon_petter_aasen}
\input{bio/carl-inge_nilsen}
\input{bio/andreas_austeng}
\fi

>>>>

% 
% \graphicspath{gfx/../}
%
\title{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}%
%
\author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
        Jon~Petter~\AA{}Åsen,~\IEEEmembership{Student Member,~IEEE,} %
        Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
        Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, 0316 Oslo, Norway.}%
\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, 7491 Trondheim, Norway.}% <-this % stops a space
% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
}%
%
% The paper headers
\markboth{IEEE Journal of Oceanic Engineering}%
{A Robust and Optimized GPU Implementation of the MVDR beamformer}%
%
% Publishers ID mark:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
%
% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}
%
% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}%
% Two limiting factors of modern phased array sonar imaging systems are the angular resolution and contrast. Adaptive beamformers target these limitatitons by weighting sensors dynamically to fit the impinging wavefield.
%
The minimum variance distortionless response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging. Unfortunately, it is very computationally complex because a spatial covariance matrix must be estimated and inverted for each image pixel. This may discourage its use unnecessarily in sonar systems which are continuously being pushed to ever higher imaging ranges and resolutions.

@@

% We benchmarked optimized implementations of the MV beamformer in MATLAB and C, running on a quad-core Intel Xeon E5620, and in CUDA running on a high-end nVidia card. The beamformers operated on data from the 32x4 element Kongsberg Martime HISAS1030 sonar, using a conservative set of robustification parameters. 
%The CPU would then be free to take on the rest of the beamforming as well.
\end{abstract}%
%
% Keywords (normally not used for peer reviews)
\begin{IEEEkeywords}
Adaptive beamforming, MVDR, Capon, sonar, active sonar, GPU, graphics processing unit, complexity.
\end{IEEEkeywords}
}%

@@

\subsection{Beamforming}
\ifPhdDoc
\begin{figure}[htbp]\centering
\includegraphics[width=\linewidth]{gfx/buske1\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[htbp]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/beamforming.eps}
\fi\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

@@

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}). This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays by
%
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
\end{align}
%
\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske2\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/implementation.eps}
\fi\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth:\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed; \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver;\newline
3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
\end{figure}
As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{finalR});
\item computing $\eRi\1$ in (\ref{weights});
\item computing the beamformer output $z$ in (\ref{finalZ}).
\end{enumerate}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

@@

If we neglect spatial and temporal averaging, then the computation of $\eR$ is reduced to a single outer product with complexity of O($M^2$), and the inversion is then of O($M^3$). This might lead us to believe that the inversion step is by far the most complex. But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$). Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
%
\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske3\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske3\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_complexity.eps}
\fi\fi%
\caption{Per-pixel computational complexity of the steps in MVDR beamforming (before any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
\end{figure}
To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}). The only excluded operation was partial pivoting used in the inversion step, but this should contribute marginally to the end result. The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$. The results are shown in \Fig{mvdr_complexity}.

@@

\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske4\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske4\figPostfix.eps}
\else
\begin{figure}[t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_build_R.eps}
\fi\fi%
\caption{Step 1: Building $\eR$. This is a visualization of how $\eR$ could be built in a case with $M=5$ sensors, with subarray size $L=3$ and temporal averaging set to $K=1$. Here $\R'_{l}$ is the sample covariance matrix for the $l$th subarray, and $\breve{\R}$ is the average of $N_K$ of these. Note that instead of performing the temporal sum last as here, one could take more temporal samples into consideration in the computation of each $r_{ij}$.}\label{mvdr_build_R}
\end{figure}

@@

\subsection[Computing the Spatial Covariance Matrix]{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}\label{computing_eR}

@@

In Fig. \ref{mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may:
%
\begin{enumerate}
\item exploit the fact that $\eR$ is Hermitian positive semidefinite to compute only one half of it;
\item avoid redundant multiplications;
\item avoid redundant summations by first computing the upper row of $\eR$ and then adding/subtracting to these iteratively to find the diagonals of $\eR$.
\end{enumerate}

\ifPhdDoc
\begin{figure}[tp]\centering
\includegraphics[width=\linewidth]{gfx/buske5\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske5\figPostfix.eps} 
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
\fi\fi%
\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig. \ref{mvdr_complexity} (higher is better) . Note how the arithmetic count is reduced by a factor 4-10 in the memory-optimized case, and by a factor 6-22 in the instruction-optimized case.}\label{mvdr_complexity_optimized}
\end{figure}

@@

\ifPhdDoc
\begin{figure}[p]\centering
\includegraphics[width=.997\linewidth]{gfx/buske6\figPostfix.eps}
% Caption is the same, but using newlines for the items in it (since it is now on a page)
\caption{MVDR implemented on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \\[.5\baselineskip]\textbf{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes interthread communication.
\\[.5\baselineskip]\textbf{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}. 
\\[.5\baselineskip]\textbf{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sums all the channels up.}\label{mvdr_implementation}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske6\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
\fi
\caption{MVDR implemented on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \emph{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes interthread communication. \emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}. \emph{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sums all the channels up.}\label{mvdr_implementation}
\fi%
\end{figure}

@@

\subsection[Inverting R]{Solving $\hat{\boldsymbol{R}}$$\;\!^{-1}$$\mathbf{1}$}

@@

\subsection[Computing z]{Computing $z$}

@@

The GPU design is summarized and compared to our CPU designs in Table \ref{implementation_summary}.
%
\begin{table}[!ht]\centering%\normalsize
% \begin{tabular}[c]{l r r r@{}  l}\hline
\begin{tabular}[l]{l >{\raggedright}p{.27\linewidth} p{.28\linewidth}}\hline
\rowcolor{tabBlue} & \bf GPU & \bf CPU \\\hline
Pixel processing & In batches & One by one \\
Threads per pixel & Multiple & One \\
1. Building $\eR$:  & & \\
\ \ \ Values computed & All & Only upper triangular \\
\ \ \ Optimization    & Minimized memory consumption & Minimized number of instructions \\
2. Computing $\eRi\1$: & & \\
\ \ \ Method & Nvidia's batch\newline Gauss Jordan solver & Custom inline\newline Cholesky solver \\
Cache utilization & Efficient & Not efficient \\
Data transfers & Minimum & Moderate
\end{tabular}
\caption{Implementation summary.}\label{implementation_summary}
\end{table}

@@

\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske7\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske7\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
\fi\fi%
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. The image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP). Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}

@@

\ifPhdDoc
\begin{figure}[p]\centering
\includegraphics[width=\linewidth]{gfx/buske8\figPostfix.eps}
% Same caption but slightly different formatting since it is now on its own page
\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.
\\[.5\baselineskip]\textbf{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in Fig. \ref{mvdr_complexity}.
\\[.5\baselineskip]\textbf{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske8\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark.eps}
\fi\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in Fig. \ref{mvdr_complexity}.\newline
\emph{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
\fi%
\end{figure}

\ifPhdDoc
\begin{figure}[p]\centering
\includegraphics[width=\linewidth]{gfx/buske9\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske9\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/code_assessment.eps}
\fi\fi%
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1 megapixel (MP) image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}

\ifPhdDoc
\begin{figure}[p]\centering
\includegraphics[width=\linewidth]{gfx/buske10\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske10\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/code_assess_flops.eps}
\fi\fi
\caption{\protect Code efficiency. An estimate of the number of floating point operations per second (Flop/s), found by dividing the theoretical complexity curves by actual run-times. This is a crude measure as it does not include any other instructions than the actual arithmetic operations in the MVDR computation.}\label{code_assess_flops}
\end{figure}

\ifPhdDoc
\begin{figure}[tbp]\centering
\includegraphics[width=\linewidth]{gfx/buske11\figPostfix.eps}
\else
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske11\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark_boston.eps}
\fi\fi
\caption{\protect MVDR benchmarks from Boston HPC centre with the new high-end Nvidia K20 Kepler GPU. The exact same scenario and code as in Fig. \ref{benchmarks} was used here. With no code alterations the performance was only improved marginally compared to running on the Quadro 6000.}\label{benchmarks_boston}
\end{figure}

@@

Even with all our efforts we were unable to obtain processing rates higher than 40\;GFlop/s on average in the desirable range of subarray sizes (\Fig{mvdr_complexity}). This is mainly due to the inversion step, since for the most part we obtained more than 100\;GFlop/s when building $\eR$. While these numbers are slightly underestimated, they are regardless far away from the Quadro 6000's peak of 1\;TFlop/s. Again, we believe this is due to the design being bound by memory bandwidth. This belief is further supported by the test results from running the code on the new K20 Kepler card, where we saw a modest factor 1.5-2 speedup. Although the Kepler card peaks at 3.5-4\;Tflop/s, its shared memory bandwidth is only approximately twice that of the Quadro 6000, hence matches the observed speedup. However, it is likely that the Kepler card would perform better had we optimized our design for it. For scientific computing, the Kepler boards are considered more attractive, with the introduction of features such as kernels calling kernels, more registers, more shared memory, and better mechanisms for hiding or removing data transfers.

@@

\ifPhdDoc
\clearpage
\appendix
% \section*{Appendix}
\renewcommand\thesection{\Roman{section}}
\else
\appendices
\fi

@@

\appendix

@@

\ifPhdDoc
   \printbibliography[title=References,heading=subbibliography]
%    \bibliographysty
%    \bibliography{library.bib}
%    \print
\else

% Originally build with bibtex:
% \bibliographystyle{IEEEtran}
% \bibliography{library.bib}

@@

\input{bio/jo_inge_buskenes}
\input{bio/jon_petter_aasen}
\input{bio/carl-inge_nilsen}
\input{bio/andreas_austeng}
\fi
